### 2016-03-22 doc/draft.Rmd
intro,methods,results AL

```{r message=F,echo=FALSE,cache=T,results='hide',warning=FALSE}
xamp_roc<-"figure 1"
Xamp_roc<-"Figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"

@@

miseq_vars<-format(miseq_vars,scientific=F)
miseq_kb<-format(miseq_kb*1000,scientific = F)

hiseq_vars<-format(hiseq_vars,scientific=F)
hiseq_kb<-format(hiseq_kb*1000,scientific = F)

@@

**Benchmarking and quality controls are required in sequence based studies of intrahost diversity.**

@@

\n
* Corresponding author

@@

The advent of next generation sequencing (NGS) is revolutionizing the study of viral evolution and molecular epidemiology through high-throughput measures of intrahost diversity. Until recently investigations of viral diversity were limited by labor intensive sequencing approaches; however, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV), and the high-throughput, cost-effective nature of NGS has lead to an explosion of studies that quantify viral diversity within and between hosts (e.g.[@Andersen:2015jj], [@Grubaugh:2015by], [@Rogers:2015ij],[@Poon:2016gf]). Given the increasing contribution of NGS to the field of virology it is important to reflect on the accuracy of these methods and the controls that are needed to ensure robust conclusions.

Because next generation sequencing of viral populations often produces tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite robust (cite). However, patient derived samples can differ in input titer by orders of magnitude (e.g. [@Lau:2013ima],[@Teunis:2015iw],[@Takeyama:2015jg]), and this variability can have drastic impacts on the sensitivity of detecting rare events. Furthermore, some library preparations result in variable sequencing coverage both within and between samples. These sources of ascertainment bias can be easily  overlooked in NGS based studies of viral diversity.

The specificity of NGS is another factor to consider when designing an experiment using patient-derived samples. Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated. As we will show, the next generation sequencing of viral populations, in particular those of RNA viruses, is error prone, and these errors must be accounted for to ensure accurate secondary analysis.

Many sample preparation pipelines have been developed to control for errors in NGS based studies of RNA virus populations; however, each approach has its own caveats that ultimately limit its application. For example, Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej]. Rolling circle cDNA synthesis produces tandem reads, generating a consensus sequence for each RNA fragment. While this method is likely to be highly sensitive for rare variant detection and can control for reverse transcription, PCR, and sequencing errors, the requirement for a large and relatively pure population of viral RNA limits its applicability to patient derived samples [@Acevedo:2014df].

"Primer Id" based methods require less input and target sequencing to the viral genome. These methods use bar-coded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  [@Jabara:2011eo]. Primer Id based methods require each bar code be physically attached to a PCR product and are most easily applied to a small targeted regions of the genome. As such, they have limited application in whole genome sequencing.

An adapted form of sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation [@Djikeng:2008cd]. In this method RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation. SISPA controls for any errors that may arise during the library amplification such as PCR biases. This method has been used in conjunction with statistical algorithms to control for accuracy in studies of intrahost influenza diversity (Ghedin group citations). However, the bar-coding reaction used in SISPA is biased and results in uneven coverage and sensitivity across the genome  [@Rosseel:2013dn].

Given the above limitations, many statistical approaches have been developed to distinguish true variants from sequencing errors [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk]. In general, these algorithms calculate base specific error rates from various metrics including but not limited to: mapping quality (MapQ), base quality (Phred), strand bias, and sequence context. True variants are identified as those with frequencies exceeding the expected error rate according to some predetermined statistical test. Despite being employed in many NGS based studies of viral diversity (e.g. ), few of these algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under conditions that mimic those found in patient-derived samples. The accuracy of such algorithms in the context of NGS studies of patient-derived viral populations is not known.

Here, we use defined populations of Influenza A and variable input titers to indirectly determine the accuracy of rare variant detection in patient-derived samples. We highlight the challenges that accompany NGS based studies of viral diversity and include a few means for improved the accuracy. This work exemplifies the controls that should be run prior to any NGS based study of viral populations and provides a comprehensive data set for benchmarking other pipelines.

## Methods

@@

Madin-Darby canine kidney cells were provided by Arnold S. Monto (University of Michigan School of Public Health) and were maintained in Dulbecco's Modified Eagle Medium (DMEM, Invitrogen) with 10% fetal bovine serum (Gibco and HyClone), 25mM HEPES (Invitogen), and 0.1875% bovine serum albumin (Life Technologies). Influenza A/WSN/33(H1N1) virus was rescued from transfected cells using the 8 plasmid reverse genetic system containing the genomic segments (pHW181-188), a kind gift from Robert Webster (St. Jude's Children's Research Hospital) [@Hoffmann:2002ih]. A biological clone of influenza A/Puerto Rico/8/1934(H1N1) was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system [@Pauly:2015eq].

@@

#### PR8 WSN33 population

For the experiment in `r Miseq`, WSN33 and PR8 viruses were plaque purified and passaged three times in MDCK cells. We then verified the sequences of the these viruses by Sanger sequencing.  Two microliters of RNA template was used  to generate cDNA in 25 $\mu\text{l}$ segment specific one step RT-PCR with $0.2\mu\text{M}$ of the following primers :   PB2-Forward-JT	(5'-GCAGGTCAATTATATTCAATATGGAAA-3'), PB2-Reverse-JT	(5'-CAAGGTCGTTTTTAAACTATTCGACAC-3'), PB1-Forward-JT (5'-GCAGGCAAACCATTTGAATGG-3'), PB1-Reverse-JT	(5'-CAAGGCATTTTTTCATGAAGGACAAG-3'), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG-3'), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG-3'), NA-Forward-JT	5'-(GCAGGAGTTTAAATGAATCCAAACC-3'), NA-Reverse-JT	(5'-CAAGGAGTTTTTTGAACAAACTACTTG-3'), HA-Forward-JT	(5'-GCAGGGGAAAATAAAAACAACCAAAAT-3'), HA-Reverse-JT	(5'-CAAGGGTGTTTTTCCTTATATTTCTGAA-3'), NP-Forward-JT	(5'-GCAGGGTAGATAATCACTCACAG-3'), NP-Reverse-JT	(5'-CAAGGGTATTTTTCTTTAATTGTCGTACT-3'), M-Forward-JT	(5'-GCAGGTAGATATTGAAAGATGAGTC-3'), M-Reverse-JT	(5'-CAAGGTAGTTTTTTACTCCAGCTCT-3'), NS-Forward-JT	(5'-GCAGGGTGACAAAGACATAATG-3'), NS-Reverse-JT	(5'-CAAAGGGTGTTTTTTATTATTAAATAAGCTG-3'). Reaction conditions were $50^\circ\text{C}$ (60 min), $94^\circ\text{C}$ (2 min), followed by 30 cycles of $94^\circ\text{C}$ (30 sec), $54^\circ\text{C}$ (30 sec), and $68^{\circ}\text{C}$ (3 min). Molar equivalents of each PCR product were pooled to generate reconstituted cDNA genomes of WSN33 and PR8. The WSN33 cDNA pool was then serially diluted into the PR8 cDNA pool to yield mixtures in which WSN33 made up 5, 2.5, 1.25, 0.63, and 0.16\% of the population. Seven hundred and fifty nanograms of purified PCR products were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator with the following settings : Intensity: 4, Duty cycle: 10%, Burst/second: 200, Duration: 80 seconds. Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L), Agencourt AMPure XP beads (Beckman Coulter A63881), and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S). Pooled libraries were sequenced on an Illumina Miseq machine using 2 x 250 paired end reads. A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments. These plasmids were mixed equimolarly and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in [@Anonymous:A25Ka2tY]. This library was prepared in identical fashion to the experimental populations and was sequenced in the same Miseq lane.

@@

Twenty point mutants were generated in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] as previously described (Visher et al., manuscript in preparation). In short, we used  overlap PCR mutagenesis to introduce the following mutations: HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T.

We passaged the 20 WSN33 point mutants and the WSN33 WT once in MDCK cells and varified the identity of the mutants by sequencing on a Illumina Miseq as above. We quantified the genome copy number of each supernant using a SuperScript III Platinum One-Step qRT-PCR (Invitrogen 11732) and univeral influenza A/B primer and probe sets [@Anonymous:tc]. Infected supernatants were mixed at equal concentrations and diluted to generate a population with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter. We diluted this mixture into WT WSN33 supernatant to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter. These 5 populations were diluted serially into basal media to generate samples with total nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter. Viral RNA was extracted from these samples and cDNA was generated in a one step multiplex RT-PCR as above. The WT WSN33 sample (10^5^ genomes per micoliter) was processed and sequenced in duplicate. We prepared libraries as before and used Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library. We pooled equal quantities (by nanogram) of each indexed library, and removed adaptor dimers using gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing an on Illumina Hiseq 2500 with 2 x 125 paired end reads. A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.

#### Duplicate populations

@@

The ability to reliably distinguish true from false single nucleotide variants (SNV) is integral to accurate NGS based studies of viral diversity. The accuracy of any SNV calling pipeline can be described in terms of its sensitivity and specificity. Sensitivity is the proportion of true variants that are properly identified; while, specificity is the proportion of true negatives that are properly identified. A closely related concept is the false positive rate, which is 1-specificity, and gives the ratio of false positive variants to true negatives. A sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.

There is an obvious trade off between sensitivity and specificity. Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity. Conversely, increased stringency can improve specificity but often reduces sensitivity. This relationship can be visualized using a receiver operating characteristic (ROC) curve (`r Xamp_roc`). An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. A variant calling pipeline must be tested against known data in order to construct an ROC curve. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real, often a p value or quality score. In a controlled benchmarking experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc`A). These points are then used to construct the curve (`r Xamp_roc`B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

###Initial accuracy

A comprehensive comparison of SNV calling approaches is beyond the scope of this work. Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving its accuracy. In doing so we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.

DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome [@Gerstung:1js]. Because it is clonal, the sequence of the control is known with a high degree of confidence, and any nonconsensus base is indicative of an error. Additionally, the control and experimental samples are processed together and are assumed to have identical noise characteristics. DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control. Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification. However, the accuracy has not been tested when applied to whole genome sequencing of a viral population amplified by reverse transcription PCR (RT-PCR).

In our first benchmarking data set we created defined mixtures of two influenza strains,WSN33 and PR8. We plaque purified and expanded WSN33 and PR8 populations. Complementary DNA from both viruses were mixed serially such that WSN33 cDNA was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq`A). These viruses differ at 491 positions (primer sites used in RT-PCR were excluded from analysis), providing 491 true positives in each dilution. On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0 [@Gerstung:1js]. Under our experimental conditions and with a slightly more stringent p value (p=0.01 vs. p=0.05) we found a moderate reduction in sensitivity  (`r miseq.sense.63` for variants at 0.63\% and `r miseq.sense.16` for variants at 0.16\%). The specificity was above 0.998 in all dilutions. While the drop in specificity appears small, a false positive rate of 0.002 corresponds to 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb` bp of the influenza genome we analyzed.

@@

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. Patient-derived populations are typically less divergent than WSN33 and PR8 (cite?), and the number of viral genomes in patient samples is much lower than in cell-culture derived populations (data not shown). To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform to account for any additional mutations that might have arisen between transfection and the passage 1 stock. Four additional mutations were found above 1% frequency (frequencies 1.2%-3.7%). We also determined the genome copy number of each stock using quantitative RT-PCR. We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A). We then serially diluted these populations into basal media to obtain mixtures with lower nucleic acid input. The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and called variants using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate to control for any mutations in the viral dilutent.

The 20 mutations present in our initial viral mixture were the only true positives considered in our analysis. Any SNV that was present in both duplicates of the wild type stock or in one of the viral clones (>1.0%) was masked and  considered neither a true positive nor a true negative.

In these populations with lower diversity and input titer, we maintained greater than 0.85 sensitivity for SNV at 1% frequency or higher. Our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C). We observed an average read depth of over 10,000 reads/bp in these samples. Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set and the fact that library preparation requires a number of sampling steps that may limit detection. During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome. The false positives outnumber the true positives by 100-fold in these populations with realistic diversity and input. We were able to increased our specificity  by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis markedly reduces sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

>>>>

```{r message=F,echo=FALSE,cache=T,results='hide',warning=FALSE}
xamp_roc<-"figure 1"
Xamp_roc<-"Figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"Figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"

@@

miseq_vars<-formatC(miseq_vars,format="d",big.mark = ",")
miseq_kb<-formatC(miseq_kb*1000,format="d",big.mark = ",")

hiseq_vars<-formatC(hiseq_vars,format="d",big.mark = ",")
hiseq_kb<-formatC(hiseq_kb*1000,format="d",big.mark = ",")

@@

**Validation and quality controls in sequence based studies of intrahost diversity**

@@

\n
*Corresponding author

@@

The advent of next generation sequencing (NGS) has enabled detailed characterization of intrahost diversity and has revolutionized studies of  viral evolution and molecular epidemiology. Until recently, investigations of viral diversity were limited by labor intensive sequencing approaches. However, it is now feasible to sequence patient-derived samples at sufficient read depth to detect rare single nucleotide variants (SNV), and the cost-effectiveness of NGS has led to an explosion of studies that quantify viral diversity within and between hosts (e.g.[@Andersen:2015jj], [@Grubaugh:2015by], [@Rogers:2015ij],[@Poon:2016gf]). Given the increasing contribution of NGS to the field of virology, it is important to reflect on the accuracy of these methods and the controls that are needed to ensure robust conclusions.

Because next generation sequencing of viral populations often produces tens of thousands of reads per genomic position, the sensitivity of these approaches has been assumed to be quite high (cite). However, patient-derived samples can differ in input titer by orders of magnitude (e.g. [@Lau:2013ima],[@Teunis:2015iw],[@Takeyama:2015jg]), and this variability can have drastic impacts on the sensitivity of detection for rare events. Furthermore, some library preparations result in variable sequencing coverage both within and between samples. These sources of ascertainment bias can be easily  overlooked in NGS based studies of viral diversity.

The specificity of NGS is another factor to consider when designing an experiment using patient-derived samples. Each sequencing platform has inherent error profiles (cite), and sample collection, target amplification, and library preparation are additional processes whereby errors can be introduced and propagated.The next generation sequencing of RNA virus populations is particularly error prone, and these errors must be accounted for to ensure accurate secondary analysis.

Many sample preparation protocols have been developed to control for errors in NGS-based studies of RNA virus populations. However, each approach has its own caveats that ultimately limit its application. Cirseq is an ingenious approach in which  template RNA is sheared and circularized prior to reverse transcription  [@Acevedo:2014ej]. Subserquent rolling circle cDNA synthesis produces tandem reads, generating a consensus sequence for each RNA fragment. While this method is likely to be highly sensitive for rare variant detection and can control for reverse transcription, PCR, and sequencing errors, the requirement for a large and relatively pure population of viral RNA limits its applicability to patient-derived samples [@Acevedo:2014df]. "Primer ID" based methods require less input and target sequencing to the viral genome. This approach relies upon barcoded primers to construct consensus sequences for each cDNA template and can control for PCR and sequencing errors  [@Jabara:2011eo]. Because Primer ID methods require taht each bar code be physically attached to a PCR product, they  are most easily applied to a small targeted regions of the genome. As such, they have limited application in whole genome sequencing.

An adapted form of sequence independent single primer amplification (SISPA) is an alternative approach that allows for whole genome sequencing and controls for errors propagated during library preparation [@Djikeng:2008cd]. In this method RT-PCR products are sheared and tagged with bar-coded random primers in a klenow reaction, prior to library preparation. SISPA controls for any errors that may arise during the library amplification such as PCR biases. This method has been used in conjunction with statistical algorithms to control for accuracy in studies of intrahost influenza diversity (Ghedin group citations). However, the bar-coding reaction used in SISPA is biased and can result in uneven coverage and sensitivity across the genome  [@Rosseel:2013dn].

Given the above limitations, many statistical approaches have been developed to distinguish true variants from sequencing errors [@Gerstung:1js; @Isakov:2015gq; @Wilm:2012br; @Macalalad:2012fx; @Koboldt:2009dk]. In general, these algorithms calculate base-specific error rates from various metrics including but not limited to: mapping quality (MapQ), base quality (Phred), strand bias, and sequence context. True variants are identified as those with frequencies exceeding the expected error rate according to some predetermined statistical test. Despite being employed in many NGS based studies of viral diversity (e.g. ), few of these algorithms have been benchmarked using defined viral populations, and to our knowledge none have been tested under conditions that mimic those found in patient-derived samples. The accuracy of such algorithms in the context of NGS studies of patient-derived viral populations is largely unknown.

Here, we use genetically-defined populations of Influenza A virus and variable input titers to indirectly determine the accuracy of rare variant detection in patient-derived samples. We highlight the challenges that accompany NGS-based studies of viral diversity and include a few means for improved the accuracy. This work exemplifies the controls that should be run prior to any NGS based study of viral populations and provides a comprehensive data set for benchmarking other pipelines.

# Methods

@@

Madin-Darby canine kidney cells were provided by Arnold S. Monto (University of Michigan School of Public Health) and were maintained in Dulbecco's Modified Eagle Medium (DMEM, Invitrogen) with 10% fetal bovine serum (Gibco and HyClone), 25mM HEPES (Invitogen), and 0.1875% bovine serum albumin (Life Technologies). Influenza A/WSN/33(H1N1) virus was rescued from transfected cells using the 8 plasmid reverse genetic system containing the genomic segments (pHW181-188), a kind gift from Robert Webster (St. Jude's Children's Research Hospital) [@Hoffmann:2002ih,@Pauly:2015eq]. A biological clone of influenza A/Puerto Rico/8/1934(H1N1) was obtained from ATCC (VR-1469) and the genomic segments were cloned into the pHW2000 reverse genetic system [@Pauly:2015eq].

@@

#### PR8-WSN33 population

For the experiment in `r Miseq`, WSN33 and PR8 viruses were plaque purified and passaged three times in MDCK cells. We then verified the sequences of the these viruses by Sanger sequencing.  Two microliters of RNA template were used  to generate cDNA in eight segment-specific one-step RT-PCR with $0.2\mu\text{M}$ of the following primers :   PB2-Forward-JT	(5'-GCAGGTCAATTATATTCAATATGGAAA-3'), PB2-Reverse-JT	(5'-CAAGGTCGTTTTTAAACTATTCGACAC-3'), PB1-Forward-JT (5'-GCAGGCAAACCATTTGAATGG-3'), PB1-Reverse-JT	(5'-CAAGGCATTTTTTCATGAAGGACAAG-3'), PA-Forward-JT	(GCAGGTACTGATTCAAAATGGAAG-3'), PA-Reverse-JT	(CAAGGTACTTTTTTGGACAGTATGG-3'), NA-Forward-JT	5'-(GCAGGAGTTTAAATGAATCCAAACC-3'), NA-Reverse-JT	(5'-CAAGGAGTTTTTTGAACAAACTACTTG-3'), HA-Forward-JT	(5'-GCAGGGGAAAATAAAAACAACCAAAAT-3'), HA-Reverse-JT	(5'-CAAGGGTGTTTTTCCTTATATTTCTGAA-3'), NP-Forward-JT	(5'-GCAGGGTAGATAATCACTCACAG-3'), NP-Reverse-JT	(5'-CAAGGGTATTTTTCTTTAATTGTCGTACT-3'), M-Forward-JT	(5'-GCAGGTAGATATTGAAAGATGAGTC-3'), M-Reverse-JT	(5'-CAAGGTAGTTTTTTACTCCAGCTCT-3'), NS-Forward-JT	(5'-GCAGGGTGACAAAGACATAATG-3'), NS-Reverse-JT	(5'-CAAAGGGTGTTTTTTATTATTAAATAAGCTG-3'). Reaction conditions were $50^\circ\text{C}$ (60 min), $94^\circ\text{C}$ (2 min), followed by 30 cycles of $94^\circ\text{C}$ (30 sec), $54^\circ\text{C}$ (30 sec), and $68^{\circ}\text{C}$ (3 min). Molar equivalents of each PCR product were pooled to generate reconstituted cDNA genomes of WSN33 and PR8. The WSN33 cDNA pool was then serially diluted into the PR8 cDNA pool to yield WSN33-PR8 mixtures in which WSN33 made up 5, 2.5, 1.25, 0.63, and 0.16\% of the population. Seven hundred and fifty nanograms of each mixture were sheared to an average size of 300-400 bp using a Covaris S220 focused ultrasonicator with the following settings : Intensity: 4, Duty cycle: 10%, Burst/second: 200, Duration: 80 seconds. Sequencing libraries were prepared from these fragmented products using the NEBNext Ultra DNA library prep kit (NEB E7370L), Agencourt AMPure XP beads (Beckman Coulter A63881), and NEBNext multiplex oligonucleotides for Illumina (NEB E7600S). Pooled libraries were sequenced on an Illumina Miseq machine using 2 x 250 paired end reads. A clonal plasmid control library was prepared from 8 plasmids containing PR8 genomic segments. These plasmids were mixed to equal molarity, and cDNA was generated using a multiplex one step RT-PCR with the primers Uni12/Inf1 (5'-GGGGGGAGCAAAAGCAGG-3'), Uni12/Inf3 (5'-GGGGGAGCGAAAGCAGG-3'), and Uni13/Inf1 (5'-5CGGGTTATTAGTAGAAACAAGG-3') as in [@Anonymous:A25Ka2tY]. This library was prepared in identical fashion to the experimental populations and was sequenced in the same Miseq lane.

@@

Twenty point mutants were generated in the WSN33 background using the pHW2000 reverse genetics system [@Hoffmann:2002ih] (Visher et al., manuscript in preparation). In short, we used  overlap PCR mutagenesis to introduce the following mutations: HA\_T1583G, HA\_G1006T, HA\_G542T, M\_T861G, M\_A541C, NA\_G1168T, NA\_C454T, NP\_A454C, NP\_A1160T, NS\_G227T, NS\_A809G, PA\_T964G, PA\_T237A, PA\_A1358T, PB1\_G599A, PB1\_G1764T, PB1\_T1288A, PB2\_A1854G, PB2\_A440T, PB2\_A1167T. Viruses were rescued from transfected cells as in [@Pauly:2015eq].

We passaged the 20 WSN33 point mutants and the WSN33 WT once in MDCK cells and verified the identity and relative clonality of the mutants by sequencing each on an Illumina Miseq as above. We quantified the genome copy number of each supernatant using a SuperScript III Platinum One-Step qRT-PCR kit (Invitrogen 11732) and universal influenza A/B primer and probe sets [@Anonymous:tc]. Equal genome equivalents of each infected supernatant were mixed  and diluted to generate a population with each mutant present at 5% frequency and a total concentration of 10^5^ genomes per microliter. We diluted this mixture into WT WSN33 supernatant to create populations in which each mutant was present at 2, 1, 0.5,and 0.2% frequency all with a total concentration of 10^5^ genomes per micoliter. These 5 populations were diluted serially into basal media to generate samples with total nucleic acid concentrations of 10^4^ and 10^3^ genomes per microliter. Viral RNA was extracted from these samples and cDNA was generated in a one step multiplex RT-PCR as above. The WT WSN33 sample (10^5^ genomes per micoliter) was processed and sequenced in duplicate. We prepared libraries as before and used Quanti PicoGreen dsDNA quantification (ThermoFisher Scientific P7589) to quantify the concentration of each indexed library. We pooled equal quantities (by nanogram) of each indexed library, and removed adaptor dimers by gel isolation with the GeneJET Gel Extraction Kit (ThermoFisher Scientific K0691) prior to sequencing an on Illumina Hiseq 2500 with 2 x 125 paired end reads. A clonal control library was processed in an identical fashion starting from an equimolar mix of 8 plasmids containing the WSN33 genomic segments.

@@

The ability to reliably identify single nucleotide variants (SNV) is integral to accurate NGS-based studies of viral diversity. The accuracy of any SNV calling pipeline can be described in terms of its sensitivity and specificity. Sensitivity is the proportion of true variants that are properly identified, and specificity is the proportion of true negatives that are properly identified. A closely related concept is the false positive rate, which is 1-specificity, and gives the ratio of false positive variants to true negatives. An assay with perfect accuracy, in which all the true variants are found and only true variants are found, has a sensitivity and specificity of 1.

There is an obvious trade-off between sensitivity and specificity. Improved sensitivity often requires less stringent criteria in variant calling, but reduces specificity. Conversely, increased stringency can improve specificity but often reduces sensitivity. This relationship can be visualized using a receiver operating characteristic (ROC) curve (`r Xamp_roc`). An ROC curve plots the sensitivity of an assay along the y-axis and 1-specificity, or the false positive rate, along the x-axis. A variant calling pipeline must be tested against known data in order to construct an ROC curve. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real, often a p value or quality score. In a controlled benchmarking experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc`A). These points are then used to construct the curve (`r Xamp_roc`B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

### Initial accuracy

A comprehensive comparison of SNV calling approaches is beyond the scope of this work. Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving its accuracy. In doing so we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is applied.

DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome [@Gerstung:1js]. Because it is clonal, the sequence of the control is known with a high degree of confidence, and any nonconsensus base is indicative of an error. Additionally, the control and experimental samples are processed together and are assumed to have identical noise characteristics. DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control. Like many variant calling algorithms, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification. However, its accuracy has not been tested when applied to whole genome sequencing of a viral population amplified by reverse transcription PCR (RT-PCR).

In our first benchmarking data set we created defined mixtures of two influenza strains,WSN33 and PR8. We plaque purified and expanded WSN33 and PR8 populations. Complementary DNA from both viruses were mixed serially such that WSN33 cDNA was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq`A). These viruses differ at 491 positions (primer sites used in RT-PCR were excluded from analysis), providing 491 true positives in each dilution. On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0 [@Gerstung:1js]. Under our experimental conditions and with a slightly more stringent p value (p=0.01 vs. p=0.05) we found a moderate reduction in sensitivity  (`r miseq.sense.63` for variants at 0.63\% and `r miseq.sense.16` for variants at 0.16\%). The specificity was above 0.998 in all dilutions. While this drop in specificity appears small, a false positive rate of 0.002 corresponds to 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb` bp of the influenza genome.

@@

Although the initial benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. Patient-derived populations are typically less divergent than WSN33 and PR8 (cite Ghedin papers), and the number of viral genomes in patient-samples is much lower than that found in cell culture. To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform to account for any additional mutations that might have arisen between transfection and the passage 1 stock. Four additional mutations were found above 1% frequency (frequencies 1.2%-3.7%). We also determined the genome copy number of each stock using quantitative RT-PCR. We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A). We then serially diluted these populations into basal media to obtain mixtures with lower nucleic acid input. The range (10^3^-10^5^ copies per microliter) matches the inputs typically found in patient-derived influenza virus samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and called variants using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate to control for any mutations in the viral dilutent.

The 20 mutations present in our initial viral mixture were the only true positives considered in our analysis. Any SNV that was present at >1% frequency in either in both duplicates of the wild type stock or any one of the viral clones was masked excluded from the analysis and  considered neither a true positive nor a true negative.

In these populations with lower diversity and input titer, we maintained greater than 0.85 sensitivity for SNV at 1% frequency or higher. . Despite a high depth of coverage (>10,000 reads per bp), our sensitivity was considerably lower for variants at or below 0.5% frequency (`r Hiseq`B and C). Therefore, the drop in sensitivity is most likely due to the 1,000-fold decrease in nucleic acid concentration compared to the first data set, and the fact that library preparation requires a number of sampling steps that may limit detection. During library preparation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with 10^5^ genomes per microliter. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome. The false positives outnumber the true positives by 100-fold in these populations with realistic diversity and input. We were able to increased our specificity  by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis markedly reduces sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.
